Computing Entropies With Nested Sampling

### 2017-06-02 paper/ms.tex
Changed to more standard notation

If an unknown quantity $x$ has a discrete probability distribution $p(x)$,
the Shannon entropy \citep{shannon} is defined as
\begin{align}
H_{x} &= -\sum_{x} p(x) \log p(x)
\end{align}
where the sum is over all of the possible values of $x$ under consideration.
The entropy quantifies the degree to which the issue
``what is the value of $x$, precisely?'' remains unresolved
\citep{knuth_questions}. For example,
if there is only one possible value which has probability 1, $H_x$ is
zero\footnote{Conventionally, $0 \times \log 0$ is defined to be equal
to $\lim_{x \to 0^+} (x\log x)$, which is zero.}.
If there are $N$ possibilities with equal probabilities $1/N$ each,
then the entropy is $H_x = \log(N)$.
If the space of possible $x$-values is continuous, the differential
entropy
\begin{align}
H_{x} &= -\int p(x) \log p(x) \, dx
\end{align}
quantifies uncertainty by generalising the log-volume
of the plausible region, when volume is defined with respect to $dx$.

@@

I prefer to write the entropy of $x$ as $H_x$ rather than $H(x)$
(or $H(X)$ as is often seen), because the latter can wrongly imply that
the entropy is a property of the value of $x$ itself, rather than
the probability distribution we have assigned to it.

@@

The entropy of the prior $p(\x)$ describes the degree to which the question
``what is the value of $\x$, precisely?'' remains unanswered, while the
entropy of the joint prior $p(\x, \d)$
describes the degree to which the question
``what is the value of the pair $(\x, \d)$, precisely?'' remains unanswered.
The degree to which the question ``what is the value of $\x$?'' would remain
unresolved if $\d$ were resolved is given by the
conditional entropy
\begin{align}
H_{\x | \d} &= - \sum_{\d} p(\d) \sum_{\x} p(\x | \d) \log p(\x | \d)
\end{align}
which is the expected value of the entropy of the posterior, averaged over
all possible datasets which might be observed.

Suppose we wanted to compute $H_{\x | \d}$, perhaps to compare it to
$H_{\x}$ and quantify how much might be learned about $\x$.
This would be difficult because the expression
for the posterior distribution
\begin{align}
p(\x | \d) &= \frac{p(\x)p(\d | \x)}{p(\d)}
\end{align}
contains the marginal likelihood integral:
\begin{align}
p(\d) &= \int p(\x) p(\d | \x) \, d\x
\end{align}
also known as the `evidence', which tends to be computable but costly.

It is important to distinguish between $H_{\x | \d}$ and the
entropy of $\x$ given a particular value of $\d$, which might be written
$H_{\x | \d=\d_0}$ (in traditional notation, $H(X|Y)$ and $H(X|Y=y)$).
The former measures the degree to which one question answers
another (ex ante), so is a function of two questions. The latter measures the
degree to which a {\em statement} answered a question (ex post),
and is a function of a question and a statement.

\subsection{Mutual information}
The mutual information is another way of describing the relevance of the
data to the parameters. Its definition, and relation to other quantities, is
\begin{align}
I_{\x, \d} &= \sum_{\x} \sum_{\d} p(\x, \d)
                       \log\left[\frac{p(\x, \d)}{p(\x)p(\d)}\right]\\
           &= H_{\x} + H_{\d} - H_{\x, \d}\\
           &= H_{\d} - H_{\d | \x}\\
           &= H_{\x} - H_{\x | \d}.
\end{align}
It can also be written in terms of the prior, likelihood, and evidence,
as
\begin{align}
I_{\x, \d} &= \sum_{\x} \sum_{\d} p(\x)p(\d | \x)
              \left[\log p(\d | \x) - \log p(\d)\right],
\end{align}
i.e., the mutual information is the prior expected value of the
log likelihood minus the log evidence.
As with the conditional entropy, we see that the difficulty appears in the
form of the log evidence,
$\log p(\d) = \log \sum_{\x} p(\x)p(\d | \x)$.

For experimental design purposes, if we want to maximise the expected amount
of information obtained from data, maximising either $I_{\x, \d}$ or
$H_{\x | \d}$ will produce the same result because the prior $p(\x)$ does not
vary with the experimental design. Reference priors
\citep{bernardo2005reference} also make use of the mutual information but in a
different way.

@@

\While{more iterations desired}
    \State $k \leftarrow 0$
            \Comment{Initialise counter}
    \State Generate $\x_{\rm ref}$ from $p(\x)$
            \Comment{Generate a reference point}
\color{green}
    \State Generate $\left\{\x_i\right\}_{i=1}^N$ from $p(\x)$
            \Comment{Generate initial NS particles}
    \State Calculate $d_i \leftarrow d(\x_i; \x_{\rm ref})$ for all $i$
            \Comment{Calculate distance of each particle from the reference
                     point}
    \State $i^* \leftarrow \textnormal{argmin}(\lambda i \rightarrow d_i)$
            \Comment{Find the worst particle (greatest distance)}
    \State $d_{\rm max} \leftarrow d_{i^*}$
            \Comment{Find the greatest distance}
    \While{$d_{\rm max} > L$}
        \State Replace $\x_{i^*}$ with $\x_{\rm new}$ from
               $p\left(\x | (d(\x) < d_{\rm max})\right)$
            \Comment{Replace worst particle}
        \State Calculate $d_{i^*} \leftarrow d(\x_{\rm new}; \x_{\rm ref})$
            \Comment{Calculate distance of new particle from reference point}
        \State $i^* \leftarrow \textnormal{argmin}(\lambda i \rightarrow d_i)$
                \Comment{Find the worst particle (greatest distance)}
        \State $d_{\rm max} \leftarrow d_{i^*}$
                \Comment{Find the greatest distance}
        \State $k \leftarrow k+1$
            \Comment{Increment counter $k$}
    \EndWhile
\color{black}
    \State $\widehat{\mathbf{h}} \leftarrow \widehat{\mathbf{h}} + [k/N]$
            \Comment{Append latest estimate to results}
\EndWhile
\State $\widehat{h_{\rm final}} \leftarrow \frac{1}{\rm num\_iterations} \sum \widehat{\mathbf{h}}$
            \Comment{Average results}
\end{algorithmic}
\caption{Calculates $H_{\x} = -\int p(\x) \log p(\x) \, d\x$
         without evaluating $\log p(\x)$. Technically, what it actually
         estimates is $\int f(\x) \int f(\x')
            \log \left[ P(d(\x', \x) < L | \x) \right]
                        \, d\x' \, d\x$, the expected value of the
        log-probability of a small region near $\x$.
        The part highlighted in green is standard Nested Sampling.
        \label{alg:algorithm}}
\end{algorithm}

@@

If we actually want the differential
entropy
\begin{align}
h_{\x} &= -\int f(x) \log f(x) \, dx
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $L$ is small, so that
\begin{align}
P(d(\x', \x) < \epsilon | \x)
    &\approx
    f(\x) \int_{d(\x', \x) < L} \, d\x'\\
    &= f(\x) \, V
\end{align}
where $V$ is the volume of the region where $d(\x', \x) < L$.

@@

This example demonstrates how the algorithm may be used to determine the
entropy for a quantity whose distribution can only be sampled.
Consider the basic statistics problem of inferring a quantity $\mu$ from
100 observations $\d = \{d_1, ..., d_{100}\}$ whose
probability distribution (conditional on $\mu$) is
\begin{align}
p(d_i | \mu) &\sim \textnormal{Normal}(\mu, 1).
\end{align}
If the prior for $\mu$ is Normal$(0, 10^2)$, then the posterior is
\begin{align}
\mu | \d &\sim \textnormal{Normal}\left(
                                       \frac{1}{100}\sum_{i=1}^{100} d_i,
                                       \left[\frac{10}{\sqrt{10001}}\right]^2
                                       \right).
\end{align}
The true value of the differential entropy $h_{\d}$ is available analytically
in this case, and its value is $146.499$. However, in most situations, the
entropy of a marginal distribution is not available in closed form.

I ran the algorithm and computed the depth using a tolerance of
$L=10^{-3}\sqrt{100}$, so that the RMS difference between
points in the two datasets was about $10^{-3}$.
From the algorithm, the average depth was
\begin{align}
\depth_{\d} &= 698.25 \pm 0.34.
\end{align}
The log-volume of a 100-dimensional ball of radius
$L=10^{-3}\sqrt{100}$ is
$-551.76$. Therefore, the differential entropy is estimated to be
\begin{align}
H_{\d} &\approx 146.49 \pm 0.34.
\end{align}
which is very close to the true value (suspiciously close, but this
was just a fluke).

@@

To convert these to differential entropies, I added the log-volumes
for 1, 101, and 102 dimensions, which are
$-6.215$, $-557.252$, and $-562.746$ respectively. This gives
the following values for the differential entropies:
\begin{align}
H'_{\tau}     &=  0.001 \pm 0.025 \label{eqn:h_tau}\\
H'_{\d}       &= -77.12 \pm 0.32   \\
H'_{\tau, \d} &= -82.79 \pm 0.31   \label{eqn:h_joint}
\end{align}
The mutual information is equal to $H_{\tau} + H_{\d} - H_{\tau, \d}$,
and can be computed from the results listed
in Equations~\ref{eqn:h_tau} to~\ref{eqn:h_joint}.
However, each of these three runs were produced with an identical random
seed for the reference points, so there is a way to reduce the uncertainty
by calculating $H_{\tau} + H_{\d} - H_{\tau, \d}$ for each reference pair.
This is an instance of the
{\em common random numbers} technique of variance reduction.
The resulting  mutual information is
$I_{\tau, \d} = 5.67 \pm 0.30$ nats.

The results for the uneven observing schedule, again in terms of log
probabilities, are:
\begin{align}
\depth_{\tau}      &= 6.216  \pm 0.025 \\
\depth_{\d}        &= 479.47 \pm 0.31  \\
\depth_{\tau, \d}  &= 480.27 \pm 0.32  \\
I_{\tau, \d}       &= 4.69 \pm 0.32
\end{align}

@@

In a Bayesian inference problem with parameters $\x$ and data $\d$, the
mutual information is
\begin{align}
I_{\x, \d} &= \iint p(\x, \d)
                        \ln \left[\frac{p(\x, \d)}{p(\x)p(\d)}\right]
                        \, d\x \, d\d \label{eqn:mutual_info2}
\end{align}
and this measures the dependence between $\x$ and $\d$. A
Monte Carlo
strategy to evaluate this is to sample from $p(\x, \d)$ and average
the value of the logarithm, which involves computing the log-evidence
\begin{align}
\ln p(\d) = \ln\left[\int p(\x)p(\d | \x) \, d\x\right]
\end{align}
for each possible data set.
Computing $p(\d)$ for even a single
data set has historically been considered difficult.

However, if we want the mutual information between the data and
{\em a subset of the parameters} (i.e., there are nuisance parameters
we don't care about), things become even more tricky. Suppose we
partition the parameters $\x$ into important parameters
$\bphi$ and nuisance parameters $\boldeta$, such that we want to calculate
the mutual information $I_{\bphi, \d}$. This is given by
\begin{align}
I_{\bphi, \d} &= \iint p(\bphi, \d)
                        \ln \left[\frac{p(\bphi, \d)}{p(\bphi)p(\d)}\right]
                        \, d\bphi \, d\d \label{eqn:mutual_info3}
\end{align}
which is equivalent to Equation~\ref{eqn:mutual_info2} but with
$\x$ replaced throughout by $\bphi$. However, this is an even more
difficult calculation than before, as $p(\d | \bphi)$, the marginal likelihood
function with the nuisance parameters integrated out, is typically
unavailable in closed form. If we were to marginalise out the nuisance
parameters $\boldeta$ explicitly, this would give us
Equation~\ref{eqn:mutual_info3} with every probability distribution
written as an explicit integral over $\boldeta$:

\begin{align}
I_{\bphi, \d} &= \iint
  \left(\int p(\bphi, \boldeta)p(\d | \bphi, \boldeta) \, d\boldeta\right)
                        \ln \left[
  \frac{\int p(\bphi, \boldeta)p(\d | \bphi, \boldeta) \, d\boldeta}
{p(\bphi)\int p(\bphi, \boldeta)p(\d | \bphi, \boldeta) \, d\bphi \, \d\boldeta}\right]
                        \, d\bphi \, d\d
\end{align}
It should not be surprising that this is costly.

>>>>

If an unknown quantity $x$ has a discrete probability distribution $p(x)$,
the Shannon entropy \citep{shannon} is defined as
\begin{align}
H(x) &= -\sum_{x} p(x) \log p(x)
\end{align}
where the sum is over all of the possible values of $x$ under consideration.
The entropy quantifies the degree to which the issue
``what is the value of $x$, precisely?'' remains unresolved
\citep{knuth_questions}. For example,
if there is only one possible value which has probability 1, $H(x)$ is
zero\footnote{Conventionally, $0 \times \log 0$ is defined to be equal
to $\lim_{x \to 0^+} (x\log x)$, which is zero.}.
If there are $N$ possibilities with equal probabilities $1/N$ each,
then the entropy is $H(x) = \log(N)$.
If the space of possible $x$-values is continuous so that $p(x)$ is a
probability density function, the differential entropy
\begin{align}
H(x) &= -\int p(x) \log p(x) \, dx
\end{align}
quantifies uncertainty by generalising the log-volume
of the plausible region, when volume is defined with respect to $dx$.

@@

Even though the entropy is written as $H(x)$, it is imperative that we
remember it is not a property of the value of $x$ itself, but a property
of the probability distribution we use to describe our ignorance of $x$.

@@

The entropy of the prior $p(\x)$ describes the degree to which the question
``what is the value of $\x$, precisely?'' remains unanswered, while the
entropy of the joint prior $p(\x, \d)$
describes the degree to which the question
``what is the value of the pair $(\x, \d)$, precisely?'' remains unanswered.
The degree to which the question ``what is the value of $\x$?'' would remain
unresolved if $\d$ were resolved is given by the
conditional entropy
\begin{align}
H(\x | \d) &= - \sum_{\d} p(\d) \sum_{\x} p(\x | \d) \log p(\x | \d)
\end{align}
which is the expected value of the entropy of the posterior, averaged over
all possible datasets which might be observed.

Suppose we wanted to compute $H(\x | \d)$, perhaps to compare it to
$H(\x)$ and quantify how much might be learned about $\x$.
This would be difficult because the expression
for the posterior distribution
\begin{align}
p(\x | \d) &= \frac{p(\x)p(\d | \x)}{p(\d)}
\end{align}
contains the marginal likelihood integral:
\begin{align}
p(\d) &= \int p(\x) p(\d | \x) \, d\x
\end{align}
also known as the `evidence', which tends to be computable but costly.

It is important to distinguish between $H(\x | \d)$ and the
entropy of $\x$ given a particular value of $\d$, which might be written
$H(\x | \d=\d_0)$.
The former measures the degree to which one question answers
another (ex ante), so is a function of two questions. The latter measures the
degree to which a {\em statement} answered a question (ex post),
and is a function of a question and a statement.

\subsection{Mutual information}
The mutual information is another way of describing the relevance of the
data to the parameters. Its definition, and relation to other quantities, is
\begin{align}
I(\x; \d) &= \sum_{\x} \sum_{\d} p(\x, \d)
                       \log\left[\frac{p(\x, \d)}{p(\x)p(\d)}\right]\\
           &= H(\x) + H(\d) - H(\x, \d)\\
           &= H(\d) - H(\d | \x)\\
           &= H(\x) - H(\x | \d).
\end{align}
It can also be written in terms of the prior, likelihood, and evidence,
as
\begin{align}
I(\x; \d) &= \sum_{\x} \sum_{\d} p(\x)p(\d | \x)
              \left[\log p(\d | \x) - \log p(\d)\right],
\end{align}
i.e., the mutual information is the prior expected value of the
log likelihood minus the log evidence.
As with the conditional entropy, we see that the difficulty appears in the
form of the log evidence,
$\log p(\d) = \log \sum_{\x} p(\x)p(\d | \x)$.

For experimental design purposes, if we want to maximise the expected amount
of information obtained from data, maximising either $I(\x; \d)$ or
$H(\x | \d)$ will produce the same result because the prior $p(\x)$ does not
vary with the experimental design. Reference priors
\citep{bernardo2005reference} also make use of the mutual information but in a
different way.

@@

\While{more iterations desired}
    \State $k \leftarrow 0$
            \Comment{Initialise counter}
    \State Generate $\x_{\rm ref}$ from $p(\x)$
            \Comment{Generate a reference point}
\color{green}
    \State Generate $\left\{\x_i\right\}_{i=1}^N$ from $p(\x)$
            \Comment{Generate initial NS particles}
    \State Calculate $d_i \leftarrow d(\x_i; \x_{\rm ref})$ for all $i$
            \Comment{Calculate distance of each particle from the reference
                     point}
    \State $i^* \leftarrow \textnormal{argmin}(\lambda i \rightarrow d_i)$
            \Comment{Find the worst particle (greatest distance)}
    \State $d_{\rm max} \leftarrow d_{i^*}$
            \Comment{Find the greatest distance}
    \While{$d_{\rm max} > L$}
        \State Replace $\x_{i^*}$ with $\x_{\rm new}$ from
               $p\left(\x | (d(\x) < d_{\rm max})\right)$
            \Comment{Replace worst particle}
        \State Calculate $d_{i^*} \leftarrow d(\x_{\rm new}; \x_{\rm ref})$
            \Comment{Calculate distance of new particle from reference point}
        \State $i^* \leftarrow \textnormal{argmin}(\lambda i \rightarrow d_i)$
                \Comment{Find the worst particle (greatest distance)}
        \State $d_{\rm max} \leftarrow d_{i^*}$
                \Comment{Find the greatest distance}
        \State $k \leftarrow k+1$
            \Comment{Increment counter $k$}
    \EndWhile
\color{black}
    \State $\widehat{\mathbf{h}} \leftarrow \widehat{\mathbf{h}} + [k/N]$
            \Comment{Append latest estimate to results}
\EndWhile
\State $\widehat{h_{\rm final}} \leftarrow \frac{1}{\rm num\_iterations} \sum \widehat{\mathbf{h}}$
            \Comment{Average results}
\end{algorithmic}
\caption{Calculates $H(\x) = -\int p(\x) \log p(\x) \, d\x$
         without evaluating $\log p(\x)$. Technically, what it actually
         estimates is $\int f(\x) \int f(\x')
            \log \left[ P(d(\x', \x) < L | \x) \right]
                        \, d\x' \, d\x$, the expected value of the
        log-probability of a small region near $\x$.
        The part highlighted in green is standard Nested Sampling.
        \label{alg:algorithm}}
\end{algorithm}

@@

If we actually want the differential
entropy
\begin{align}
H(\x) &= -\int f(x) \log f(x) \, dx
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $L$ is small, so that
\begin{align}
P(d(\x', \x) < \epsilon | \x)
    &\approx
    f(\x) \int_{d(\x', \x) < L} \, d\x'\\
    &= f(\x) \, V
\end{align}
where $V$ is the volume of the region where $d(\x', \x) < L$.

@@

This example demonstrates how the algorithm may be used to determine the
entropy for a quantity whose distribution can only be sampled.
Consider the basic statistics problem of inferring a quantity $\mu$ from
100 observations $\d = \{d_1, ..., d_{100}\}$ whose
probability distribution (conditional on $\mu$) is
\begin{align}
p(d_i | \mu) &\sim \textnormal{Normal}(\mu, 1).
\end{align}
If the prior for $\mu$ is Normal$(0, 10^2)$, then the posterior is
\begin{align}
\mu | \d &\sim \textnormal{Normal}\left(
                                       \frac{1}{100}\sum_{i=1}^{100} d_i,
                                       \left[\frac{10}{\sqrt{10001}}\right]^2
                                       \right).
\end{align}
The true value of the differential entropy $H(\d)$ is available analytically
in this case, and its value is $146.499$. However, in most situations, the
entropy of a marginal distribution is not available in closed form.

I ran the algorithm and computed the depth using a tolerance of
$L=10^{-3}\sqrt{100}$, so that the RMS difference between
points in the two datasets was about $10^{-3}$.
From the algorithm, the average depth was
\begin{align}
\depth_{\d} &= 698.25 \pm 0.34.
\end{align}
The log-volume of a 100-dimensional ball of radius
$L=10^{-3}\sqrt{100}$ is
$-551.76$. Therefore, the differential entropy is estimated to be
\begin{align}
H(\d) &\approx 146.49 \pm 0.34.
\end{align}
which is very close to the true value (suspiciously close, but this
was just a fluke).

@@

To convert these to differential entropies, I added the log-volumes
for 1, 101, and 102 dimensions, which are
$-6.215$, $-557.252$, and $-562.746$ respectively. This gives
the following values for the differential entropies:
\begin{align}
H(\tau)     &=  0.001 \pm 0.025 \label{eqn:h_tau}\\
H(\d)       &= -77.12 \pm 0.32   \\
H(\tau, \d) &= -82.79 \pm 0.31   \label{eqn:h_joint}
\end{align}
The mutual information is equal to $H(\tau) + H(\d) - H(\tau, \d)$,
and can be computed from the results listed
in Equations~\ref{eqn:h_tau} to~\ref{eqn:h_joint}.
However, each of these three runs were produced with an identical random
seed for the reference points, so there is a way to reduce the uncertainty
by calculating $H(\tau) + H(\d) - H(\tau, \d)$ for each reference pair.
This is an instance of the
{\em common random numbers} technique of variance reduction.
The resulting  mutual information is
$I(\tau; \d) = 5.67 \pm 0.30$ nats.

The results for the uneven observing schedule, again in terms of log
probabilities, are:
\begin{align}
\depth_{\tau}      &= 6.216  \pm 0.025 \\
\depth_{\d}        &= 479.47 \pm 0.31  \\
\depth_{\tau, \d}  &= 480.27 \pm 0.32  \\
I(\tau; \d)        &= 4.69 \pm 0.32
\end{align}

@@

In a Bayesian inference problem with parameters $\x$ and data $\d$, the
mutual information is
\begin{align}
I(\x; \d) &= \iint p(\x, \d)
                        \ln \left[\frac{p(\x, \d)}{p(\x)p(\d)}\right]
                        \, d\x \, d\d \label{eqn:mutual_info2}
\end{align}
and this measures the dependence between $\x$ and $\d$. A
Monte Carlo
strategy to evaluate this is to sample from $p(\x, \d)$ and average
the value of the logarithm, which involves computing the log-evidence
\begin{align}
\ln p(\d) = \ln\left[\int p(\x)p(\d | \x) \, d\x\right]
\end{align}
for each possible data set.
Computing $p(\d)$ for even a single
data set has historically been considered difficult.

However, if we want the mutual information between the data and
{\em a subset of the parameters} (i.e., there are nuisance parameters
we don't care about), things become even more tricky. Suppose we
partition the parameters $\x$ into important parameters
$\bphi$ and nuisance parameters $\boldeta$, such that we want to calculate
the mutual information $I(\bphi; \d)$. This is given by
\begin{align}
I(\bphi; \d) &= \iint p(\bphi, \d)
                        \ln \left[\frac{p(\bphi, \d)}{p(\bphi)p(\d)}\right]
                        \, d\bphi \, d\d \label{eqn:mutual_info3}
\end{align}
which is equivalent to Equation~\ref{eqn:mutual_info2} but with
$\x$ replaced throughout by $\bphi$. However, this is an even more
difficult calculation than before, as $p(\d | \bphi)$, the marginal likelihood
function with the nuisance parameters integrated out, is typically
unavailable in closed form. If we were to marginalise out the nuisance
parameters $\boldeta$ explicitly, this would give us
Equation~\ref{eqn:mutual_info3} with every probability distribution
written as an explicit integral over $\boldeta$:

\begin{align}
I(\bphi; \d) &= \iint
  \left(\int p(\bphi, \boldeta)p(\d | \bphi, \boldeta) \, d\boldeta\right)
                        \ln \left[
  \frac{\int p(\bphi, \boldeta)p(\d | \bphi, \boldeta) \, d\boldeta}
{p(\bphi)\int p(\bphi, \boldeta)p(\d | \bphi, \boldeta) \, d\bphi \, \d\boldeta}\right]
                        \, d\bphi \, d\d
\end{align}
It should not be surprising that this is costly.
